A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-10 doc/popl2013/langdef.tex
Patched version of theorem 2.

\newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
\newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
\newcommand{\dplus}{\text{+\!+}}
\newcommand{\llbracket}{[\![}
\newcommand{\rrbracket}{]\!]}
\newcommand{\tuple}[2]{\langle #1, #2 \rangle}
\makeatletter
% macros for consistency flavors
\newcommand{\foo}[1]{\cancel{#1}}
\newcommand{\simH}{\sim_{\!H}}
% usually: \mapconsist{\Omega}
\newcommand{\mapconsist}[2][M]{#1 \foo\sim #2}
% usually: \localconsist{L}{\Gamma}
\newcommand{\localconsist}[3][M]{#2 \foo\simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
% usually: \storeconsist{S}{H}
\newcommand{\storeconsist}[2]{#1 \foo\sim #2}
% usually: \valueconsist{v}{T}
\newcommand{\valueconsist}[3][M]{#2 \foo\simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
% usually: \traceconsist{E}
\newcommand{\traceconsist}[2][H]{#2 \foo\sim #1}
% usually: \privconsist{E}{\Phi}
\newcommand{\privconsist}[3][M]{#2 \@ifmtarg{#1}{:}{:_{\!#1}} #3}

@@

The first option is to serialize the constituent traces in program
order.  Although each constituent trace has been checked for coherence, the
checking of some operations may have been deferred due to non-empty
clobber sets.  It is possible that when an interleaving of these traces
is attemped in the parent expression's context, there may no valid interleaving
if the wrong guesses were made in the derivation of a subexpression.
However, we can show that as long as any valid interleaving
exists, the serialization of the constituent traces is valid.
To prove this result, we use the following lemma, the proof
of which follows directly from the definition of $apply$.

\begin{lem}
\label{lemma:applychain}
\rm
For any initial store $S$ and memory traces $E_1$ and $E_2$,
$$apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2).$$
\end{lem}

\begin{thm}
\label{thm:sequential}
\rm
Let $S_0$ be a store, $C$ a clobber set and $\oton{E}$ memory traces to be
interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$, in program order.  If there exists any $E''$ that is a
valid interleaving of $\oton{E}$, then $E'$ is also valid interleaving of $\oton{E}$.
\end{thm}

% FIX ME FIX ME FIX ME
\begin{proof}
To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the four conditions:
\begin{enumerate}
\item $any\_interleave(E', \oton{E})$ is proven by induction on
the length of $E_1$, followed by $E_2$, and so on.
\item $coherent(S_0, C, E')$ is shown by considering each constituent trace's subsequence in $E'$
in order, using Lemma~\ref{lemma:applychain} with the coherence of each $E_i$.
\item $seq\_equiv(S_0, C, E', \oton{E})$ follows trivially by matching the structure of $E'$
to the corresponding  pieces $\oton{E}$.
\item $coherent(S_0, C, E')$ is shown using the sequential equivalence of $E''$.  The value 
returned by each read in $E''$ is equal to the value in the store that results
from the application of all previous constituent traces and the operations before the read in its
own trace.  This is exactly the sequence of preceding operations of the corresponding read in $E'$,
and Lemma~\ref{lemma:applychain} allows us to transform the chain of $apply$ operations into a 
single $apply$ operation on the whole sequence, satisfying the coherence condition for the read
in $E'$.  This property applies for all reads in $E''$, and since $E'$ and $E''$ are
interleavings of the same traces, they have the same reads, so all reads
in $E'$ are coherent.
\end{enumerate}
\end{proof}

>>>>

\newcommand{\oton}[1]{{#1}_1,\ldots,{#1}_n}
\newcommand{\otok}[2]{{#2}_1,\ldots,{#2}_{#1}}
\newcommand{\dplus}{\text{+\!+}}
\newcommand{\llbracket}{[\![}
\newcommand{\rrbracket}{]\!]}
\newcommand{\tuple}[2]{\langle #1, #2 \rangle}
\makeatletter
% macros for consistency flavors
\newcommand{\simH}{\sim_{\!H}}
% usually: \mapconsist{\Omega}
\newcommand{\mapconsist}[2][M]{#1 \sim #2}
% usually: \localconsist{L}{\Gamma}
\newcommand{\localconsist}[3][M]{#2 \simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
% usually: \storeconsist{S}{H}
\newcommand{\storeconsist}[2]{#1 \sim #2}
% usually: \valueconsist{v}{T}
\newcommand{\valueconsist}[3][M]{#2 \simH \@ifmtarg{#1}{#3}{#1 \llbracket #3 \rrbracket}}
% usually: \traceconsist{E}
\newcommand{\traceconsist}[2][H]{#2 \sim #1}
% usually: \privconsist{E}{\Phi}
\newcommand{\privconsist}[3][M]{#2 \@ifmtarg{#1}{:}{:_{\!#1}} #3}

@@

The first option is to serialize the evaluation of the subexpressions that
make up an expression and attempt to generate a valid interleaving by simply
concatenating the constituent traces in program order.  However, there is a
complication related to the nondeterminism that is allowed for reads.  Although
each subexpression's trace will have been checked for coherence, the checking
of some operations may have been deferred due to non-empty
clobber sets.  It is possible that when an interleaving of these traces
is attemped in the parent expression's context, there may no valid interleaving
if the wrong guesses were made in the derivation of a subexpression.
The solution in the case of serial execution is to not expand the clobber
set for subexpressions, instead using the parent expression's clobber set.

Only a single rule is modified for this alternate execution - $e_{n+1}$ in
{\sc E-Call} uses the original $C$ instead of $C'$.  Note that this change
can result in there being no terminating execution for an expression in
which the only terminating executions required a particular (non-program-)
ordering of subexpression evaluation, but the Legion core language does not
guarantee parallel execution in any case, so the evaluation of such an 
expression is unlikely to terminate even under the original clobber set.
We can, however, show that any terminating execution that results from using
the restricted clobber set is a valid execution under the original:

\begin{lem}
\label{lem:clobberreduce}
\rm 
Let $\typeenv{e}{T}$ and $\opsenv{e}{v}{E}$ and $\storeconsist{S}{H}$.  If $C \subseteq C'$, then
$\opsenv[C=C']{e}{v}{E}$.
\end{lem}

We now show that the concatenation of the memory traces from these serially
executed subexpression is always a valid interleaving:

\begin{thm}
\label{thm:sequential}
\rm
Let $\oton{e}$ be expressions such that $\opsenv[S={S_{i-1}}]{e_i}{v_i}{E_i}$, 
where $S_i = apply(S_{i-1}, E_i)$.  If $E' = E_1 \dplus \ldots \dplus E_n$, then
$valid\_interleave(S_0, C, E', \oton{E})$.
\end{thm}

\begin{proof}
First, we must show that $any\_interleave(E', \oton{E})$.  This is done by
induction on the length of $E_1$, followed by $E_2$, and so on.  Next, we must
show that $E'$ is coherent for both $L_{excl}$ and $L_{atomic}$.  This is done by
matching the structure of $E'$ with $\oton{E}$ and the use of the following
lemma: $apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2)$.  With the
question of coherence addressed, the sequential equivalence of $E'$ is trivial
for both the exclusive and atomic cases.
\end{proof}
